Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-19 Multiresolution_framework/ltx/manuscript.tex
added appendix for the lower bound

 and the posterior distribution of the field over time. Here $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the entire sequence of observations $\mathbf Y$ parameterised  by $\boldsymbol\theta$ known as the likelihood function. When maximisation of the likelihood function is difficult to achieve directly the EM algorithm can be used to find increasingly tighter lower bounds on the likelihood of the parameters so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. By exploiting the total probability rule it is rather straightforward to show that 

@@

is a lower bound on the log-likelihood function \cite{Bishop2006} where $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ is the joint probabilty distribution of the states and the observations. Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function.... The rest of this section is to construct the lower bound $\mathcal Q$ for the linear state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}. The joint probability distribution $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ can be written as

@@

The authors acknowledge that there is, and will always be, a discrepancy between the model and cortex. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge. Additional avenues for future work include extending the model to capture a $2^{nd}$-order synaptic response kernel, and time delays according to spatial resolutions. The estimation techniques should also be extended to deal with spatiotemporal heterogeneity in the kernel. Finally, future work should be directed towards applying and validating the framework on real data.
\section{Appendix}
\subsection{Inner product of two B-spline scaling functions}\label{ap:InnerProductOfBsplines}
In this section, an analytic formula is derived for the inner product of two B-spline scaling functions. Consider two B-spline scaling functions of order $m$ and $m'$. The support of $N_m\left(s\right)$ is $\left[ 0,m\right]$ and  is symmetric with respect to $r=\frac{m}{2}$, i.e.
\begin{align}\label{eq:APP-SymmetrySplieEq}
 N_{m}\left(\frac{m}{2}+r\right)=N_{m}\left(\frac{m}{2}-r\right).
\end{align}
A direct consequence of \eqref{eq:APP-SymmetrySplieEq} is 
\begin{align}
 N_{m}\left(r\right)=N_{m}\left(m-r\right).
\end{align}
Therefore
\begin{align}
\int_{-\infty}^{+\infty}N_{m}\left(r-l_{1}\right)N_{m'}\left(r-l_{2}\right)ds&=\int_{-\infty}^{+\infty}N_{m}\left(m-r+l_{1}\right)N_{m'}\left(r-l_{2}\right)dr \nonumber \\
&=\int_{-\infty}^{+\infty}N_{m}\left(m+l_{1}-l_{2}-u\right)N_{m'}\left(u\right)du \nonumber \\
&=\left(N_m \ast N_{m'}\right) \left(m+l_{1}-l_{2}\right) \nonumber \\
&=N_{m+m'}\left(m+l_{1}-l_{2}\right) \nonumber \\
&=N_{m+m'}\left(m'+l_{2}-l_{1}\right).
\end{align}
\subsection{Induction}\label{ap:Induction}
For $n_1$ in equation \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} we have
\begin{align}
 N_{m}\left(r\right)&=\sum_{n_1=0}^{m} \left[\mathbf p\right]_{n_1} N_{m}\left(2r-n_1\right) \label{eq:App-TwoScalepair1},
  \end{align}
which clearly holds (see \eqref{eq:MRA-TwoScalepair1}). Now assuming \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} for $n_{j-1}$ we have
\begin{align}\label{eq:app-ConvertingFormulaForscalingFunctions}
 &N_m(r)=\nonumber \\
&\sum_{n_1,n_2, \dots n_{j-1}=0}^{m}\left[\mathbf p\right]_{n_1} \left[\mathbf p\right]_{n_2}\dots \left[\mathbf p\right]_{n_{j-1}}N_m(2^{j-1}r-k_{n_1,n_2, \dots, n_{j-1}}),
\end{align}
$N_m(2^{j-1}r-k_{n_1,n_2, \dots, n_{j-1}})$ can be written in terms of $N_m(2^{j}r)$ as
\begin{align}\label{eq:AppendixNj-1intermsofNj}
 &N_m(2^{j-1}r-k_{n_1,n_2, \dots, n_{j-1}})=\nonumber \\
&\sum_{n_j=0}^m \left[\mathbf p\right]_{n_j}N_m(2^jr-2\times k_{n_1,n_2, \dots, n_{j-1}}-n_j)\nonumber \\
&=\sum_{n_j=0}^m \left[\mathbf p\right]_{n_j}N_m(2^jr- k_{n_1,n_2, \dots, n_{j}})
\end{align}
Substituting \eqref{eq:AppendixNj-1intermsofNj} in to \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} completes the proof.
% \cut{Future work should be directed towards extending the framework to account for a second-order synaptic response kernel, heterogeneity in the connectivity and perhaps time delays at lower spatial resolutions. Most importantly, future work should be directed towards applying and validating the framework on real data.}

% use section* for acknowledgement
%\section*{Acknowledgment}

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

>>>>

 and the posterior distribution of the field over time. Here $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the entire sequence of observations $\mathbf Y$ parameterised  by $\boldsymbol\theta$ known as the likelihood function. The goal of the EM algorithm is to maximise  the likelihood function in the presence of the hidden variables, the weights of the field basis functions in this case. The EM finds increasingly tighter lower bounds on the likelihood of the parameters so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. By exploiting the total probability rule it is rather straightforward to show that 

@@

is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix...)  where $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ is the joint probability distribution of the states and the observations. .... The rest of this section is to construct the lower bound $\mathcal Q$ for the linear state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}. The joint probability distribution $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ can be written as

@@

The authors acknowledge that there is, and will always be, a discrepancy between the model and cortex. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge. Additional avenues for future work include extending the model to capture a $2^{nd}$-order synaptic response kernel, and time delays according to spatial resolutions. The estimation techniques should also be extended to deal with spatiotemporal heterogeneity in the kernel. Finally, future work should be directed towards applying and validating the framework on real data.
\section{Appendix}
\subsection{Inner product of two B-spline scaling functions}\label{ap:InnerProductOfBsplines}
In this section, an analytic formula is derived for the inner product of two B-spline scaling functions. Consider two B-spline scaling functions of order $m$ and $m'$. The support of $N_m\left(s\right)$ is $\left[ 0,m\right]$ and  is symmetric with respect to $r=\frac{m}{2}$, i.e.
\begin{align}\label{eq:APP-SymmetrySplieEq}
 N_{m}\left(\frac{m}{2}+r\right)=N_{m}\left(\frac{m}{2}-r\right).
A direct consequence of \eqref{eq:APP-SymmetrySplieEq} is 
 N_{m}\left(r\right)=N_{m}\left(m-r\right).
Therefore
\int_{-\infty}^{+\infty}N_{m}\left(r-l_{1}\right)N_{m'}\left(r-l_{2}\right)ds&=\int_{-\infty}^{+\infty}N_{m}\left(m-r+l_{1}\right)N_{m'}\left(r-l_{2}\right)dr \nonumber \\
&=\int_{-\infty}^{+\infty}N_{m}\left(m+l_{1}-l_{2}-u\right)N_{m'}\left(u\right)du \nonumber \\
&=\left(N_m \ast N_{m'}\right) \left(m+l_{1}-l_{2}\right) \nonumber \\
&=N_{m+m'}\left(m+l_{1}-l_{2}\right) \nonumber \\
&=N_{m+m'}\left(m'+l_{2}-l_{1}\right).
\subsection{Induction}\label{ap:Induction}
For $n_1$ in equation \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} we have
 N_{m}\left(r\right)&=\sum_{n_1=0}^{m} \left[\mathbf p\right]_{n_1} N_{m}\left(2r-n_1\right) \label{eq:App-TwoScalepair1},
which clearly holds (see \eqref{eq:MRA-TwoScalepair1}). Now assuming \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} for $n_{j-1}$ we have
\begin{align}\label{eq:app-ConvertingFormulaForscalingFunctions}
&\sum_{n_1,n_2, \dots n_{j-1}=0}^{m}\left[\mathbf p\right]_{n_1} \left[\mathbf p\right]_{n_2}\dots \left[\mathbf p\right]_{n_{j-1}}N_m(2^{j-1}r-k_{n_1,n_2, \dots, n_{j-1}}),
$N_m(2^{j-1}r-k_{n_1,n_2, \dots, n_{j-1}})$ can be written in terms of $N_m(2^{j}r)$ as
\begin{align}\label{eq:AppendixNj-1intermsofNj}
 &N_m(2^{j-1}r-k_{n_1,n_2, \dots, n_{j-1}})=\nonumber \\
&\sum_{n_j=0}^m \left[\mathbf p\right]_{n_j}N_m(2^jr-2\times k_{n_1,n_2, \dots, n_{j-1}}-n_j)\nonumber \\
&=\sum_{n_j=0}^m \left[\mathbf p\right]_{n_j}N_m(2^jr- k_{n_1,n_2, \dots, n_{j}})
Substituting \eqref{eq:AppendixNj-1intermsofNj} in to \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} completes the proof. 
\subsection{Lower bound on likelihood function}
A derivation of the lower bound that is based on \cite{Bishop2006} is provided in this Appendix. Let the entire sequence of observations is denoted by $\mathbf Y$, and the sequence of hidden variables by $\mathbf X$ and the parameter set by $\boldsymbol\Theta$. Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. By exploiting the  total probability rule the following expression can be written \cite{Koralov2007}
 \ln~p(\mathbf Y;\boldsymbol\Theta)=\ln\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\Theta)d\mathbf X.
For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y\mid\boldsymbol\Theta)=\mathcal{L}(q,\boldsymbol\Theta)+\mathrm{KL}(q\parallel p),
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\Theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)}{(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\Theta)$ in \eqref{eq:EM-decomposition1} after noting that $q(\mathbf X)$ is a normalised distribution that integrates to one. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$ and $q(\mathbf X)$ and as a distance it is always non-negative, $\mathrm{KL}(q\parallel p)\ge 0$, with equality if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. It therefore follows from \eqref{eq:EM-loglikelihooddecomposition} that $\mathcal{L}(q,\boldsymbol\Theta)\le \ln~p(\mathbf Y\mid\boldsymbol\Theta)$, in other words \eqref{eq:EM-decomposition1} is a lower bound on the log-likelihood function. The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\Theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\Theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\Theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\Theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the Kullback-Liebler distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. At this point, the lower bound becomes equal to the log-likelihood function. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\Theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\Theta,\boldsymbol\Theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
where $ \mathbf E_{\boldsymbol \Theta'}\left[ .\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \Theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\Theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood.

\ifCLASSOPTIONcaptionsoff
  \newpage
\fi
